---
title: "Bulk RNA-seq Salmon + DESeq2 Standard Analysis"
author: "Nacev Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This R Markdown template is the Nacev Lab standard downstream analysis for gene-level bulk RNA-seq.

The standard input is Salmon `quant.sf` output, imported with `tximport`, followed by DESeq2 differential expression analysis. This template uses fake example sample names and placeholder paths. Before running it on a real project, replace the settings and sample metadata in the setup sections below.

This template is Salmon-based, not STAR-count-based. STAR or featureCounts outputs may be useful for special projects, but routine gene-level bulk RNA-seq should start from Salmon quantification results.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# ---- User edits required -----------------------------------------------------
# Replace these placeholders with paths from the project workspace.
# Do not point to this template repository for raw data or project results.

project_dir <- "/path/to/project"
salmon_dir <- file.path(project_dir, "salmon")
outdir <- file.path(project_dir, "results", "bulk_rna_seq_deseq2")
gtf_path <- "/path/to/reference/gencode.annotation.gtf"

# Current lab standards:
#   human: hg38
#   mouse: mm10
organism <- "human"      # Change to "mouse" for mouse projects.
genome_build <- "hg38"   # Use "mm10" for mouse projects.

# Edit this formula to match the approved experimental design.
# Examples:
#   ~ condition
#   ~ batch + condition
#   ~ genotype + treatment
design_formula <- ~ condition

# Differential expression thresholds used for result summaries and plots.
alpha <- 0.05
lfc_threshold <- 1
min_count <- 10
min_samples_per_group <- 2

dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(outdir, "tables"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(outdir, "plots"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(outdir, "objects"), showWarnings = FALSE, recursive = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tximport)
  library(readr)
  library(DESeq2)
  library(ggplot2)
  library(pheatmap)
  library(rtracklayer)
  library(clusterProfiler)
  library(msigdbr)
  library(enrichplot)
  library(AnnotationDbi)
  library(matrixStats)
})

if (organism == "human") {
  suppressPackageStartupMessages(library(org.Hs.eg.db))
  OrgDb <- org.Hs.eg.db
  msig_species <- "Homo sapiens"
  kegg_organism <- "hsa"
} else if (organism == "mouse") {
  suppressPackageStartupMessages(library(org.Mm.eg.db))
  OrgDb <- org.Mm.eg.db
  msig_species <- "Mus musculus"
  kegg_organism <- "mmu"
} else {
  stop("Set organism to either 'human' or 'mouse'.")
}
```

```{r helper-functions}
save_plot <- function(plot_obj, filename, width = 8, height = 6) {
  ggsave(file.path(outdir, "plots", filename), plot_obj, width = width, height = height)
}

clean_ensembl <- function(x) {
  sub("\\..*", "", x)
}

map_gene_symbols <- function(ensembl_ids) {
  clean_ids <- clean_ensembl(ensembl_ids)
  mapped <- AnnotationDbi::select(
    OrgDb,
    keys = unique(clean_ids),
    keytype = "ENSEMBL",
    columns = c("SYMBOL", "ENTREZID")
  )
  mapped <- mapped[!duplicated(mapped$ENSEMBL), ]
  data.frame(
    gene_id = ensembl_ids,
    ensembl = clean_ids,
    symbol = mapped$SYMBOL[match(clean_ids, mapped$ENSEMBL)],
    entrez = mapped$ENTREZID[match(clean_ids, mapped$ENSEMBL)],
    stringsAsFactors = FALSE
  )
}

result_to_table <- function(res_obj) {
  df <- as.data.frame(res_obj)
  df$gene_id <- rownames(df)
  symbols <- map_gene_symbols(df$gene_id)
  df$ensembl <- symbols$ensembl[match(df$gene_id, symbols$gene_id)]
  df$symbol <- symbols$symbol[match(df$gene_id, symbols$gene_id)]
  df$entrez <- symbols$entrez[match(df$gene_id, symbols$gene_id)]
  df$direction <- ifelse(
    is.na(df$padj) | df$padj >= alpha | abs(df$log2FoldChange) < lfc_threshold,
    "not_significant",
    ifelse(df$log2FoldChange > 0, "up", "down")
  )
  df[order(df$padj), c(
    "gene_id", "ensembl", "symbol", "entrez", "baseMean", "log2FoldChange",
    "lfcSE", "stat", "pvalue", "padj", "direction"
  )]
}
```

# 2. Sample Metadata

Replace this fake example table with a project-specific metadata file. The `sample_id` values should match the Salmon output folders, and each folder should contain `quant.sf`.

```{r sample-metadata}
# Option A, recommended for real projects:
# sample_table <- read.csv(file.path(project_dir, "metadata", "sample_sheet.csv"))

# Option B, fake teaching example:
sample_table <- data.frame(
  sample_id = c(
    "celllineA_control_rep1",
    "celllineA_control_rep2",
    "celllineA_control_rep3",
    "celllineA_treated_rep1",
    "celllineA_treated_rep2",
    "celllineA_treated_rep3"
  ),
  condition = factor(
    c("control", "control", "control", "treated", "treated", "treated"),
    levels = c("control", "treated")
  ),
  replicate = c("rep1", "rep2", "rep3", "rep1", "rep2", "rep3"),
  stringsAsFactors = FALSE
)

sample_table$files <- file.path(salmon_dir, sample_table$sample_id, "quant.sf")
rownames(sample_table) <- sample_table$sample_id

# Check that the files exist before continuing.
missing_files <- sample_table$files[!file.exists(sample_table$files)]
if (length(missing_files) > 0) {
  stop(
    "Missing Salmon quant.sf files:\n",
    paste(missing_files, collapse = "\n"),
    "\n\nEdit salmon_dir or sample_table before running the analysis."
  )
}

sample_table
```

```{r contrasts}
# Define planned comparisons before looking at results.
# Direction matters: numerator vs denominator means treated / control here.
contrast_table <- data.frame(
  contrast_id = "treated_vs_control",
  variable = "condition",
  numerator = "treated",
  denominator = "control",
  stringsAsFactors = FALSE
)

contrast_table
```

# 3. Import Salmon Results With tximport

This section follows the Salmon import strategy used in the previous Salmon-based reference notebook. The GTF must match the Salmon transcriptome index.

```{r tx2gene}
gtf <- rtracklayer::import(gtf_path)

tx2gene <- data.frame(
  tx_id = mcols(gtf)$transcript_id,
  gene_id = mcols(gtf)$gene_id,
  stringsAsFactors = FALSE
)
tx2gene <- unique(na.omit(tx2gene))

head(tx2gene)
```

```{r tximport}
files <- sample_table$files
names(files) <- sample_table$sample_id

txi <- tximport(
  files,
  type = "salmon",
  tx2gene = tx2gene,
  ignoreAfterBar = TRUE
)

head(txi$counts)
```

# 4. Create DESeq2 Object

```{r deseq-object}
dds <- DESeqDataSetFromTximport(
  txi = txi,
  colData = sample_table,
  design = design_formula
)

dds
```

# 5. Filter Low-Count Genes

This default filter keeps genes with at least `r min_count` counts in at least `r min_samples_per_group` samples. For complex designs, review whether a group-aware filter is more appropriate.

```{r filtering}
raw_counts <- counts(dds)

keep <- rowSums(raw_counts >= min_count) >= min_samples_per_group
dds <- dds[keep, ]

cat("Genes before filtering:", nrow(raw_counts), "\n")
cat("Genes after filtering:", nrow(dds), "\n")
```

# 6. Run DESeq2

```{r run-deseq}
dds <- DESeq(dds)

saveRDS(dds, file.path(outdir, "objects", "deseq2_object.rds"))
```

# 7. Export Differential Expression Results

```{r results}
all_results <- list()
all_result_tables <- list()
all_sig_genes <- character()

for (i in seq_len(nrow(contrast_table))) {
  contrast_id <- contrast_table$contrast_id[i]
  variable <- contrast_table$variable[i]
  numerator <- contrast_table$numerator[i]
  denominator <- contrast_table$denominator[i]

  res <- results(
    dds,
    contrast = c(variable, numerator, denominator),
    alpha = alpha
  )

  all_results[[contrast_id]] <- res
  res_table <- result_to_table(res)
  all_result_tables[[contrast_id]] <- res_table

  sig_table <- res_table[
    !is.na(res_table$padj) &
      res_table$padj < alpha &
      abs(res_table$log2FoldChange) >= lfc_threshold,
  ]

  all_sig_genes <- union(all_sig_genes, sig_table$gene_id)

  write.csv(
    res_table,
    file.path(outdir, "tables", paste0("DESeq2_", contrast_id, "_all_genes.csv")),
    row.names = FALSE
  )
  write.csv(
    sig_table,
    file.path(outdir, "tables", paste0("DESeq2_", contrast_id, "_significant_genes.csv")),
    row.names = FALSE
  )

  cat("\n", contrast_id, "\n", sep = "")
  print(summary(res))
  cat("Significant genes:", nrow(sig_table), "\n")
}
```

# 8. Variance-Stabilized Expression

```{r vst}
vst_data <- vst(dds, blind = TRUE)
vst_matrix <- assay(vst_data)

saveRDS(vst_data, file.path(outdir, "objects", "vst_object.rds"))
write.csv(vst_matrix, file.path(outdir, "tables", "vst_expression_matrix.csv"))
```

# 9. PCA

```{r pca}
pca_data <- plotPCA(vst_data, intgroup = "condition", returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

p_pca <- ggplot(pca_data, aes(x = PC1, y = PC2, color = condition, label = name)) +
  geom_point(size = 4) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  theme_minimal(base_size = 12) +
  labs(
    title = "PCA of bulk RNA-seq samples",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance")
  )

print(p_pca)
save_plot(p_pca, "PCA_PC1_PC2.pdf", width = 8, height = 6)
```

# 10. Sample Distance Heatmap

```{r sample-distance}
sample_dist <- dist(t(vst_matrix))
sample_dist_mat <- as.matrix(sample_dist)
rownames(sample_dist_mat) <- sample_table$sample_id
colnames(sample_dist_mat) <- sample_table$sample_id

annotation_col <- as.data.frame(colData(dds)[, c("condition"), drop = FALSE])

pdf(file.path(outdir, "plots", "Sample_distance_heatmap.pdf"), width = 8, height = 7)
pheatmap(
  sample_dist_mat,
  clustering_distance_rows = sample_dist,
  clustering_distance_cols = sample_dist,
  annotation_col = annotation_col,
  main = "Sample-to-sample distances"
)
dev.off()

pheatmap(
  sample_dist_mat,
  clustering_distance_rows = sample_dist,
  clustering_distance_cols = sample_dist,
  annotation_col = annotation_col,
  main = "Sample-to-sample distances"
)
```

# 11. Volcano Plots

```{r volcano}
plot_volcano <- function(res_table, contrast_id, annotate_top_n = 10) {
  plot_df <- res_table
  plot_df$neg_log10_padj <- -log10(plot_df$padj)
  plot_df$significance <- "not_significant"
  plot_df$significance[
    !is.na(plot_df$padj) &
      plot_df$padj < alpha &
      plot_df$log2FoldChange >= lfc_threshold
  ] <- "up"
  plot_df$significance[
    !is.na(plot_df$padj) &
      plot_df$padj < alpha &
      plot_df$log2FoldChange <= -lfc_threshold
  ] <- "down"

  label_df <- plot_df[
    plot_df$significance != "not_significant" & !is.na(plot_df$symbol),
  ]
  label_df <- head(label_df[order(label_df$padj), ], annotate_top_n)

  p <- ggplot(plot_df, aes(x = log2FoldChange, y = neg_log10_padj, color = significance)) +
    geom_point(alpha = 0.6, size = 1.4, na.rm = TRUE) +
    geom_vline(xintercept = c(-lfc_threshold, lfc_threshold), linetype = "dashed") +
    geom_hline(yintercept = -log10(alpha), linetype = "dashed") +
    geom_text(
      data = label_df,
      aes(label = symbol),
      color = "black",
      size = 3,
      vjust = -0.6,
      check_overlap = TRUE
    ) +
    scale_color_manual(values = c(
      "not_significant" = "grey70",
      "up" = "#D55E00",
      "down" = "#0072B2"
    )) +
    theme_minimal(base_size = 12) +
    labs(
      title = paste("Volcano plot:", contrast_id),
      x = "log2 fold change",
      y = "-log10 adjusted p-value",
      color = ""
    )

  print(p)
  save_plot(p, paste0("Volcano_", contrast_id, ".pdf"), width = 8, height = 7)
}

for (contrast_id in names(all_result_tables)) {
  plot_volcano(all_result_tables[[contrast_id]], contrast_id)
}
```

# 12. Gene Expression Boxplots

Use this section to inspect individual genes of interest across groups. Replace `genes_to_plot` with project-relevant gene symbols after confirming they are present in the dataset.

```{r gene-boxplots}
genes_to_plot <- c("GENE1", "GENE2")

symbol_map <- map_gene_symbols(rownames(vst_matrix))
plot_expr <- data.frame(
  gene_id = rep(rownames(vst_matrix), ncol(vst_matrix)),
  sample_id = rep(colnames(vst_matrix), each = nrow(vst_matrix)),
  vst_expression = as.vector(vst_matrix),
  stringsAsFactors = FALSE
)
plot_expr$symbol <- symbol_map$symbol[match(plot_expr$gene_id, symbol_map$gene_id)]
plot_expr <- merge(plot_expr, sample_table, by = "sample_id")

for (gene_symbol in genes_to_plot) {
  gene_df <- plot_expr[plot_expr$symbol == gene_symbol, ]
  if (nrow(gene_df) == 0) {
    message("Skipping ", gene_symbol, ": not found in mapped gene symbols.")
    next
  }

  p <- ggplot(gene_df, aes(x = condition, y = vst_expression, fill = condition)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.15, size = 2) +
    theme_minimal(base_size = 12) +
    labs(
      title = paste("Expression of", gene_symbol),
      x = "Condition",
      y = "VST expression"
    ) +
    theme(legend.position = "none")

  print(p)
  save_plot(p, paste0("Expression_boxplot_", gene_symbol, ".pdf"), width = 5, height = 5)
}
```

# 13. Heatmaps

```{r heatmap-top-variable}
top_variable_n <- min(500, nrow(vst_matrix))
top_variable <- head(order(matrixStats::rowVars(vst_matrix), decreasing = TRUE), top_variable_n)
top_variable_mat <- vst_matrix[top_variable, ]
top_variable_symbols <- map_gene_symbols(rownames(top_variable_mat))$symbol
rownames(top_variable_mat) <- ifelse(
  is.na(top_variable_symbols),
  rownames(top_variable_mat),
  top_variable_symbols
)

pdf(file.path(outdir, "plots", "Heatmap_top_variable_genes.pdf"), width = 9, height = 10)
pheatmap(
  top_variable_mat,
  scale = "row",
  annotation_col = annotation_col,
  show_rownames = FALSE,
  main = paste("Top", top_variable_n, "variable genes")
)
dev.off()
```

```{r heatmap-significant}
if (length(all_sig_genes) >= 2) {
  sig_mat <- vst_matrix[intersect(all_sig_genes, rownames(vst_matrix)), , drop = FALSE]
  sig_symbols <- map_gene_symbols(rownames(sig_mat))$symbol
  rownames(sig_mat) <- ifelse(is.na(sig_symbols), rownames(sig_mat), sig_symbols)

  pdf(file.path(outdir, "plots", "Heatmap_significant_genes_union.pdf"), width = 9, height = 12)
  pheatmap(
    sig_mat,
    scale = "row",
    annotation_col = annotation_col,
    show_rownames = nrow(sig_mat) <= 75,
    main = "Union of significant genes"
  )
  dev.off()
} else {
  message("Skipping significant-gene heatmap: fewer than 2 significant genes.")
}
```

# 14. Enrichment Analysis

This section follows the downstream enrichment style from the previous analysis reference. It performs GO over-representation analysis and Hallmark GSEA for each contrast. KEGG GSEA is included as an optional block because KEGG availability can vary by organism and network access.

```{r enrichment-helpers}
make_entrez_list <- function(res_table, direction = c("up", "down", "all")) {
  direction <- match.arg(direction)
  df <- res_table[
    !is.na(res_table$padj) &
      res_table$padj < alpha &
      abs(res_table$log2FoldChange) >= lfc_threshold,
  ]
  if (direction == "up") df <- df[df$log2FoldChange > 0, ]
  if (direction == "down") df <- df[df$log2FoldChange < 0, ]
  unique(na.omit(df$entrez))
}

make_ranked_list <- function(res_table) {
  df <- res_table[!is.na(res_table$stat) & !is.na(res_table$entrez), ]
  ranks <- df$stat
  names(ranks) <- df$entrez
  ranks <- sort(ranks, decreasing = TRUE)
  ranks[!duplicated(names(ranks))]
}

plot_if_not_empty <- function(enrich_obj, title, filename) {
  enrich_df <- as.data.frame(enrich_obj)
  if (is.null(enrich_df) || nrow(enrich_df) == 0) {
    message("No enrichment terms for ", title)
    return(invisible(NULL))
  }
  p <- dotplot(enrich_obj, showCategory = 15, font.size = 8) + ggtitle(title)
  print(p)
  save_plot(p, filename, width = 10, height = 8)
}
```

```{r go-ora}
for (contrast_id in names(all_result_tables)) {
  res_table <- all_result_tables[[contrast_id]]
  universe_entrez <- unique(na.omit(res_table$entrez))

  genes_all <- make_entrez_list(res_table, "all")
  genes_up <- make_entrez_list(res_table, "up")
  genes_down <- make_entrez_list(res_table, "down")

  go_all <- enrichGO(
    gene = genes_all,
    universe = universe_entrez,
    OrgDb = OrgDb,
    keyType = "ENTREZID",
    ont = "BP",
    pAdjustMethod = "BH",
    readable = TRUE
  )
  go_up <- enrichGO(
    gene = genes_up,
    universe = universe_entrez,
    OrgDb = OrgDb,
    keyType = "ENTREZID",
    ont = "BP",
    pAdjustMethod = "BH",
    readable = TRUE
  )
  go_down <- enrichGO(
    gene = genes_down,
    universe = universe_entrez,
    OrgDb = OrgDb,
    keyType = "ENTREZID",
    ont = "BP",
    pAdjustMethod = "BH",
    readable = TRUE
  )

  write.csv(as.data.frame(go_all), file.path(outdir, "tables", paste0("GO_BP_", contrast_id, "_all.csv")), row.names = FALSE)
  write.csv(as.data.frame(go_up), file.path(outdir, "tables", paste0("GO_BP_", contrast_id, "_up.csv")), row.names = FALSE)
  write.csv(as.data.frame(go_down), file.path(outdir, "tables", paste0("GO_BP_", contrast_id, "_down.csv")), row.names = FALSE)

  plot_if_not_empty(go_all, paste("GO BP:", contrast_id, "all DE genes"), paste0("GO_BP_", contrast_id, "_all.pdf"))
  plot_if_not_empty(go_up, paste("GO BP:", contrast_id, "upregulated genes"), paste0("GO_BP_", contrast_id, "_up.pdf"))
  plot_if_not_empty(go_down, paste("GO BP:", contrast_id, "downregulated genes"), paste0("GO_BP_", contrast_id, "_down.pdf"))
}
```

```{r hallmark-gsea}
hallmark <- msigdbr(species = msig_species, category = "H")
hallmark_t2g <- unique(hallmark[, c("gs_name", "entrez_gene")])

for (contrast_id in names(all_result_tables)) {
  ranked_list <- make_ranked_list(all_result_tables[[contrast_id]])

  if (length(ranked_list) < 20) {
    message("Skipping Hallmark GSEA for ", contrast_id, ": ranked list is too short.")
    next
  }

  gsea_hallmark <- GSEA(
    ranked_list,
    TERM2GENE = hallmark_t2g,
    pvalueCutoff = 0.05,
    pAdjustMethod = "BH",
    minGSSize = 15,
    maxGSSize = 500,
    verbose = FALSE
  )

  gsea_df <- as.data.frame(gsea_hallmark)
  write.csv(gsea_df, file.path(outdir, "tables", paste0("GSEA_Hallmark_", contrast_id, ".csv")), row.names = FALSE)

  if (nrow(gsea_df) > 0) {
    gsea_hallmark@result$Description <- gsub("HALLMARK_", "", gsea_hallmark@result$Description)
    gsea_hallmark@result$Description <- gsub("_", " ", gsea_hallmark@result$Description)
    p <- dotplot(gsea_hallmark, showCategory = 15, split = ".sign", font.size = 8) +
      facet_grid(. ~ .sign) +
      ggtitle(paste("Hallmark GSEA:", contrast_id))
    print(p)
    save_plot(p, paste0("GSEA_Hallmark_", contrast_id, "_dotplot.pdf"), width = 12, height = 8)
  }
}
```

```{r optional-kegg-gsea, eval=FALSE}
# Optional: enable this chunk when KEGG access and organism settings are validated.
for (contrast_id in names(all_result_tables)) {
  ranked_list <- make_ranked_list(all_result_tables[[contrast_id]])
  gsea_kegg <- gseKEGG(
    geneList = ranked_list,
    organism = kegg_organism,
    pvalueCutoff = 0.05,
    pAdjustMethod = "BH",
    minGSSize = 15,
    maxGSSize = 500,
    verbose = FALSE
  )
  write.csv(as.data.frame(gsea_kegg), file.path(outdir, "tables", paste0("GSEA_KEGG_", contrast_id, ".csv")), row.names = FALSE)
}
```

# 15. Save Session Information

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(outdir, "sessionInfo.txt"))
sessionInfo()
```
